Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-09-21 Manuscript/00-main_word.Rmd
Add word version

---
title: Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data
author:
- name: Boyi Guo*
  num: a
- name: Byron C. Jaeger
  num: b 
- name: AKM Fazlur Rahman
  num: a   
- name: D. Leann Long
  num: a  
- name: Nengjun Yi*
  num: a
address:
- num: a
  org: Department of Biostatistics, University of Alabama at Birmingham, Birmingham, USA
- num: b
  org: Department of Biostatistics and Data Science, Wake Forest University, Winston-Salem, USA
corres: "Boyi Guo and Nengjun Yi, Department of Biostatistics, University of Alabama at Birmingham, Birmingham, USA. \\email{boyiguo1@uab.edu} and  \\email{nyi@uab.edu}"
presentaddress: This is sample for present address text this is sample for present address text
authormark: Guo et al.
articletype: Research article
received: 2021-01-01
revised: 2021-02-01
accepted: 2021-03-01
abstract: "There are proposals that extend the classical generalized additive models (GAMs) to accommodate the high-dimensional (p>>n) problem using group penalties and priors. These models rarely smooth the nonlinear effect properly because of the sparse regularization. It leads to excess shrinkage on coefficients and inaccurate prediction. Besides, most of these GAMs consider an “all-in-all-out” approach for function selection, rendering them difficult to answer if nonlinear effects are necessary. While some Bayesian models can address the shortcomings, using Markov chain Monte Carlo (MCMC) algorithms for model fitting creates a new challenge, scalability. Hence, we propose Bayesian hierarchical generalized additive models as a solution. For smooth and sparse solutions, the proposed models employ bi-part spike-and-slab spline priors, including mixture double exponential and mixture normal priors. Two optimization-based algorithms, EM-Coordinate Descent and EM-Iterative Weighted Least Squares, are developed for expedited computation. Simulation studies and metabolomics data analyses demonstrate improved performance against state-of-art models, mgcv, COSSO and sparse Bayesian GAM. The software implementation of the proposed models is freely available via an R package BHAM."
keywords: Spike-and-Slab Priors; High-Dimensional Data; Generalized Additive Models; EM-IWLS; EM-Coordinate Decent
bibliography: bibfile.bib
output: 
  rticles::sim_article:
    keep_tex: TRUE
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
---

@@

>>>>

---
title: "Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data"

abstract: "There are proposals that extend the classical generalized additive models (GAMs) to accommodate the high-dimensional (p>>n) problem using group penalties and priors. These models rarely smooth the nonlinear effect properly because of the sparse regularization. It leads to excess shrinkage on coefficients and inaccurate prediction. Besides, most of these GAMs consider an “all-in-all-out” approach for function selection, rendering them difficult to answer if nonlinear effects are necessary. While some Bayesian models can address the shortcomings, using Markov chain Monte Carlo (MCMC) algorithms for model fitting creates a new challenge, scalability. Hence, we propose Bayesian hierarchical generalized additive models as a solution. For smooth and sparse solutions, the proposed models employ bi-part spike-and-slab spline priors, including mixture double exponential and mixture normal priors. Two optimization-based algorithms, EM-Coordinate Descent and EM-Iterative Weighted Least Squares, are developed for expedited computation. Simulation studies and metabolomics data analyses demonstrate improved performance against state-of-art models, mgcv, COSSO and sparse Bayesian GAM. The software implementation of the proposed models is freely available via an R package BHAM."
keywords: Spike-and-Slab Priors; High-Dimensional Data; Generalized Additive Models; EM-IWLS; EM-Coordinate Decent
bibliography: bibfile.bib
output: 
  word_document:
    reference_docx: template.docx
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
---

@@

\newpage

### 2021-09-21 Manuscript/02-Method.Rmd
Add word version

Meanwhile, following the classical IWLS, we can approximate the generalized model likelihood at each iteration with a weighted normal likelihood:
$$
  f(\textbf{y}|\bs \beta, \phi) \approx MVN(\textbf{z}|X\beta, \Sigma w_i^{-1}\phi)
$$
where the ‘normal response’ $z_i$ and ‘weight’ $w_i$ are called the pseudo-response and pseudo-weight respectively.The pseudo data and the pseudo-weight are calculated by
$$
\begin{aligned}
z_i &= \hat\eta_i - \frac{L^{'}(y_i|\hat\eta_i)}{L^{''}(y_i|\hat\eta_i)}& w_i &= - L^{''}(y_i|\hat\eta_i),
\end{aligned}
$$
where $L^{'}(y_i|\hat\eta_i, \hat \phi)$ and $L^{''}(y_i|\hat\eta_i, \hat \phi)$ are the first and second derivative of the log density, $\log f(\textbf{y}_i|\bs \beta, \phi)$. With $z\sim N(X\beta, \Sigma\phi)$ and $\beta \sim N(0, \phi\Sigma_{\tau^2})$, we can augment the two normal distributions and iterative solve for $\bs \beta$ and $\phi$.
  To note, as the EM-IWLS algorithm is based on the IWLS algorithm, an Hessian matrix can be derived. The Hessian matrix can be further used to construct variance-covariance matrix of coefficient estimates and in turn used for statistical inferences.

>>>>

Meanwhile, following the classical IWLS, we can approximate the generalized model likelihood at each iteration with a weighted normal likelihood:
$$
  f(\textbf{y}|\bs \beta, \phi) \approx MVN(\textbf{z}|X\beta, \Sigma w_i^{-1}\phi)
$$
where the ‘normal response’ $z_i$ and ‘weight’ $w_i$ are called the pseudo-response and pseudo-weight respectively.The pseudo data and the pseudo-weight are calculated by
$$
\begin{aligned}
z_i &= \hat\eta_i - \frac{L^{'}(y_i|\hat\eta_i)}{L^{''}(y_i|\hat\eta_i)}& w_i &= - L^{''}(y_i|\hat\eta_i),
\end{aligned}
$$
where $L^{'}(y_i|\hat\eta_i, \hat \phi)$ and $L^{''}(y_i|\hat\eta_i, \hat \phi)$ are the first and second derivative of the log density, $\log f(\textbf{y}_i|\bs \beta, \phi)$. With $z\sim N(X\beta, \Sigma\phi)$ and $\beta \sim N(0, \phi\Sigma_{\tau^2})$, we can augment the two normal distributions and derive the estimates for $\bs \beta$ and $\phi$ in each step in closed form equations. To note, an estimate of the variance-covariance matrix can be derived in the EM-IWLS algorithm and in turn can be used for statistical inferences.

### 2021-09-21 Manuscript/SS_GAM.tex
Add word version

Similar to the EM-CD algorithm, the EM-IWLS algorithm is an iterative
EM-based algorithm where iterative weighted least square algorithm are
used to find the estimate of \(\boldsymbol{\beta}, \phi\) that maximizes
\(E(Q_1)\), instead of the coordinate descent algorithm. It also can fit
the models with a wider range of priors, e.g.~mixture \(t\) prior,
including mixture normal prior as an special case, and mixture double
exponential prior. The basic idea is to formulate the priors as
hierarchical normal distributions and augment with the weighted normal
likelihood of the generalized model, such that the model is feasible to
fit via least squares algorithms. \citep{Yi2012} The primary purpose of
the EM algorithm here is mitigate the computational burden associate
with the hierarchical normal formulation.

A double exponential prior, \(\beta|S \sim DE(0, S)\) can be formulated
as a hierarchical normal prior with unknown variances, following gamma
distributions, integrated out: \begin{align*}
  \beta &\sim N(0, \tau^2)\\
  \tau^2 & \sim Gamma(1, 1/(2S^2)), 
\end{align*} where \(S = (1-\gamma)s_0 + \gamma s_1\) for spike-and-slab
lasso priors. Now \(Q_1\) needs to account for the hyper prior of
\(\tau^2\) and is expressed as \begin{equation}\label{eq:Q1_IWLS}
Q_1(\boldsymbol{\beta}, \phi) = \log f(\textbf{y}|\boldsymbol{\beta}, \phi) + \log f(\phi) + \sum\limits_{j=1}^p\left[\log f(\beta^0_j|{\tau^0}^2_j) + \log f({\tau^0}^2_j| S_j^0)+\sum\limits_{k=1}^{K_j} \{\log f(\beta^{*}_{jk}|{\tau^{*}}^2_{jk})+\log f({\tau^*}^2_{jk}| S^*_j)\}\right].
\end{equation} As \(\boldsymbol{\tau}^2\) are not of primary interest,
we treat them as the ``missing'' data besides the latent indicators
\(\boldsymbol{\gamma}\), and hence need to construct the expectation
\(E_{\boldsymbol{\gamma}, \boldsymbol{\tau}^2|\Theta^{(t-1)}}(Q_1)\). To
note, unlike the coefficients of the non-linear terms \(\beta^*_{jk}\)
which share the same latent indicator \(\gamma^*_j\), \(\tau^2_{jk}\) is
coefficient specific for \(\beta^*_{jk}\). Primarily, four quantities
need to be calculate to formulate
\(E(Q_1), E({S_j^0}^{-1}|\beta_j^0, s_0, s_1), E(S^{-1}_j|\boldsymbol{\beta}_j^*, s_0, s_1), E({\tau^0}^2_{j}|S_j^0, \beta^0_j) \text{ and } E({\tau^*}^2_{jk}|S_j^*, \beta^*_{jk})\).
As neither \(E({S_j^0}^{-1}|\beta_j^0, s_0, s_1)\) nor
\(E(S^{-1}_j|\boldsymbol{\beta}_j^*, s_0, s_1)\) depends on \(\tau^2\)s,
they can be derived via Bayes' theorem as in Equation
(\ref{eq:exp_scale}). Shown in Yi and Ma \citep{Yi2012}, the conditional
posterior density of \(\tau^{-2}\) is an inverse Gaussian distribution,
and hence \begin{align*}
 E({\tau^0}^{-2}_{j}|S_j^0, \beta^0_j) ={S_j^0}^{-1}/|\beta^0_j| \qquad E({\tau^*}^{-2}_{jk}|S_j^*, \beta^*_{jk})={S_j^*}^{-1}/|\beta^*_{jk}|,
\end{align*} where \(S_j^0\) and \(S_j^*\) are replaced by the
expectation.

To simplify \(E(Q_1)\) by removing constant additive terms, we have
\begin{equation}\label{eq:Q1_IWLS}
E(Q_1(\boldsymbol{\beta}, \phi)) = \log f(\textbf{y}|\boldsymbol{\beta}, \phi) + \log f(\phi) - \sum\limits_{j=1}^p\left[ \frac{({\beta_j^0}^2)E({\tau_j^0}^{-2}}{2\phi} +\sum\limits_{k=1}^{K_j} \frac{({\beta_{jk}^*}^2)E({\tau_{jk}^*}^{-2}}{2\phi}\right],
\end{equation} where
\(\frac{({\beta_{jk}^*}^2)E({\tau_{jk}^*}^{-2}}{2\phi}\) is the kernel
of a normal density with mean 0 and variance
\(E({\tau_{jk}^*}^{-2})\phi\).

Meanwhile, following the classical IWLS, we can approximate the
generalized model likelihood at each iteration of the algorithms by the
weighted normal likelihood: \[
  f(\textbf{y}_i|\boldsymbol{\beta}, \phi) \approx Normal(\textbf{z}_i|X_i\beta, w_i^{-1}\phi)
\] where the `normal response' \(z_i\) and `weight' \(w_i\) are called
the pseudo-response and pseudo-weight respectively.The pseudo data and
the pseudo-weight are calculated by \[
\begin{aligned}
z_i &= \hat\eta_i - \frac{L^{'}(y_i|\hat\eta_i)}{L^{''}(y_i|\hat\eta_i)}& w_i &= - L^{''}(y_i|\hat\eta_i),
\end{aligned}
\] where \(L^{'}(y_i|\hat\eta_i, \hat \phi)\) and
\(L^{''}(y_i|\hat\eta_i, \hat \phi)\) are the first and second
derivative of the log density,
\(\log f(\textbf{y}_i|\boldsymbol{\beta}, \phi)\). With
\(z\sim N(X\beta, \Sigma\phi)\) and
\(\beta \sim N(0, \phi\Sigma_{\tau^2})\), we can augment the two normal
distributions and iterative solve for \(\boldsymbol{\beta}\) and
\(\phi\). To note, as the EM-IWLS algorithm is based on the IWLS
algorithm, an Hessian matrix can be derived. The Hessian matrix can be
further used to construct variance-covariance matrix of coefficient
estimates and in turn used for statistical inferences.

>>>>

Similar to the EM-CD algorithm, the EM-IWLS algorithm is an iterative
EM-based algorithm where the iterative weighted least squares algorithm
is used to find the estimate of \(\boldsymbol{\beta}, \phi\) that
maximizes \(E(Q_1)\). The iterative weighted least squares algorithm was
originally proposed to fit the classical generalized linear models, and
generalized to fit some Bayesian hierarchical models.\citep{Gelman2013}
Yi and Ma \citep{Yi2012} formulated Student's t-distribution and double
exponential distribution as hierarchical normal distributions such that
generalized linear models with shrinkage priors can be easily fitted
using IWLS in combination with EM algorithm. In this work, we adapt the
EM-IWLS paradigm to fit BHAM with spike-and-slab spline prior, with the
full potential to adopt Student's t, double exponential and mixture
Student's t priors when sparse assumption is not necessary.

A double exponential prior, \(\beta|S \sim DE(0, S)\) can be formulated
as a hierarchical normal prior with unknown variance
\(\tau^2 \sim Gamma(1, 1/(2S^2))\) integrated out: \begin{align*}
  \beta &\sim N(0, \tau^2)\\
  \tau^2 & \sim Gamma(1, 1/(2S^2)), 
\end{align*} For the mixture double exponential priors, we can define
the scale parameter \(S = (1-\gamma)s_0 + \gamma s_1\) following
Equation (\ref{eq:ssl}). The change in the prior formulation in turns
leads to the change in the log posterior density function, as \(Q_1\)
needs to account for the hyper prior of \(\tau^2\):
\begin{equation}\label{eq:Q1_IWLS}
Q_1(\boldsymbol{\beta}, \phi) = \log f(\textbf{y}|\boldsymbol{\beta}, \phi) + \log f(\phi) + \sum\limits_{j=1}^p\left[\log f(\beta^0_j|{\tau^0}^2_j) + \log f({\tau^0}^2_j| S_j^0)+\sum\limits_{k=1}^{K_j} \{\log f(\beta^{*}_{jk}|{\tau^{*}}^2_{jk})+\log f({\tau^*}^2_{jk}| S^*_j)\}\right].
\end{equation} Since \(\boldsymbol{\tau}^2\) are not of our primary
interest, we treat them as the ``missing'' data in addition to the
latent indicators \(\boldsymbol{\gamma}\), and hence construct the
expectation
\(E_{\boldsymbol{\gamma}, \boldsymbol{\tau}^2|\Theta^{(t-1)}}(Q_1)\) in
the E-step. To note, unlike the same latent indicator \(\gamma^*_j\)
which is shared by the coefficients of the non-linear terms
\(\beta^*_{jk}\) for \(k = 1, \dots, K_j\) , \(\tau^2_{jk}\) is
coefficient specific for \(\beta^*_{jk}\).
\(E({S_j^0}^{-1}|\beta_j^0, s_0, s_1), E(S^{-1}_j|\boldsymbol{\beta}_j^*, s_0, s_1), E({\tau^0}^2_{j}|S_j^0, \beta^0_j) \text{ and } E({\tau^*}^2_{jk}|S_j^*, \beta^*_{jk})\)
needs to be calculated to formulate \(E(Q_1)\). As neither
\(E({S_j^0}^{-1}|\beta_j^0, s_0, s_1)\) nor
\(E(S^{-1}_j|\boldsymbol{\beta}_j^*, s_0, s_1)\) depends on \(\tau^2\)s,
they can be derived using Equation (\ref{eq:exp_scale}) via Bayes'
Theorem. On the other hand, \(\tau^{2}\) following a gamma distribution
is a conjugate prior for the normal variance, and the conditional
posterior density of \(\tau^{-2}\) is an inverse Gaussian distribution.
\(E({\tau^0}^{-2}_{j})\) and \(E({\tau^*}^{-2}_{jk})\) are calculated
using the closed form equation \begin{align*}
 E({\tau^0}^{-2}_{j}|S_j^0, \beta^0_j) ={S_j^0}^{-1}/|\beta^0_j| \qquad E({\tau^*}^{-2}_{jk}|S_j^*, \beta^*_{jk})={S_j^*}^{-1}/|\beta^*_{jk}|,
\end{align*} where \(S_j^0\) and \(S_j^*\) are replaced by the
expectation and \(\beta\)s are replaced with \(\beta^{(t-1)}\). With
simplification (up to constant additive terms), we have
\begin{equation}\label{eq:EQ1_IWLS}
E(Q_1) = \log f(\textbf{y}|\boldsymbol{\beta}, \phi) + \log f(\phi) - \sum\limits_{j=1}^p\left[ \frac{({\beta_j^0}^2)}{2\phi E{\tau_j^0}^{2}} +\sum\limits_{k=1}^{K_j} \frac{{\beta_{jk}^*}^2}{2\phi E({\tau_{jk}^*}^{2})}\right].
\end{equation} \(\frac{\beta^2}{2\phi E(\tau^{2})}\) can be seen as the
kernel of a normal density with mean 0 and variance
\(\phi E(\tau^{2})\).

Meanwhile, following the classical IWLS, we can approximate the
generalized model likelihood at each iteration with a weighted normal
likelihood: \[
  f(\textbf{y}|\boldsymbol{\beta}, \phi) \approx MVN(\textbf{z}|X\beta, \Sigma w_i^{-1}\phi)
\] where the `normal response' \(z_i\) and `weight' \(w_i\) are called
the pseudo-response and pseudo-weight respectively.The pseudo data and
the pseudo-weight are calculated by \[
\begin{aligned}
z_i &= \hat\eta_i - \frac{L^{'}(y_i|\hat\eta_i)}{L^{''}(y_i|\hat\eta_i)}& w_i &= - L^{''}(y_i|\hat\eta_i),
\end{aligned}
\] where \(L^{'}(y_i|\hat\eta_i, \hat \phi)\) and
\(L^{''}(y_i|\hat\eta_i, \hat \phi)\) are the first and second
derivative of the log density,
\(\log f(\textbf{y}_i|\boldsymbol{\beta}, \phi)\). With
\(z\sim N(X\beta, \Sigma\phi)\) and
\(\beta \sim N(0, \phi\Sigma_{\tau^2})\), we can augment the two normal
distributions and derive the estimates for \(\boldsymbol{\beta}\) and
\(\phi\) in each step in closed form equations. To note, an estimate of
the variance-covariance matrix can be derived in the EM-IWLS algorithm
and in turn can be used for statistical inferences.
